\chapter*{General Introduction}
\addcontentsline{toc}{chapter}{General Introduction}
\markboth{General Introduction}{General Introduction}
\label{chap:introduction}


The dramatic advancement in natural language processing in the last years owes much to the Transformer architecture, which can represent long and complex dependencies in a parallelized manner and has been able to achieve state-of-the-art accuracy in almost every single sequence modeling problem. This expressive capability, however, does not come for free because training large language models of this complexity comes with high memory requirements, huge amounts of computational throughput and high energy consumption, establishing a steep performance ceiling which makes deploying them in production or in a resource-limited setting virtually infeasible.

The issue of closing the gap between the scale used for training and that for deployment has led to the proliferation of neural network pruning as a top compression technique. The ultimate goal of pruning is to detect and eliminate either redundant parameters or structures from the already trained or during the training of network and not alter its current performance much. Pruning is, in fact, not a single problem but a highly complex optimization. One has to consider not only the architecture of the pruning, but also the condition for pruning (Whether to rely on weights magnitudes, on gradients or loss curvature), and the condition needed to make recovery.

As the literature on Transformer pruning has expanded, a critical limitation has become apparent across the state-of-the-art. The vast majority of current methods rely on local scoring, which means evaluating the importance of a single weight, attention head, or layer in isolation before the joint effect of all removals is known. While local criteria are highly effective at moderate compression levels, they degrade rapidly at high sparsity. At aggressive compression ratios, the global allocation of the sparsity budget across the model's layers becomes vastly more important than the precision of any individual local score. Evaluating local criteria without considering global structural interactions limits the true potential of Transformer compression.

The objective of this report is to review comprehensively the state-of-the-art in Transformer pruning, analyze the existing methodologies, and formally isolate this \("\)allocation problem\("\) to motivate future search-based pruning frameworks.

To achieve this, the report is divided into two main parts. The first part establishes the foundational knowledge required for this study and spans two chapters.Chapter \ref{chap:foundations} covers the theoretical foundations of neural networks. Chapter \ref{chap:specialized_architectures} then examines the architecture of the Transformer, defines its structural components and the specific hardware-aware efficiency metrics that practically drive the need for compression.

The second part surveys the extensive literature on Transformer pruning and evaluates the state of the art. It begins with Chapter \ref{chap:neural_network_pruning}, which formalizes the pruning problem mathematically by categorizing the different structural regimes. Chapter 4 reviews Unstructured and Heterogeneous Search-based methods, focusing on techniques that prioritize theoretical flexibility and search space exploration at the weight and layer-profile levels. Chapter 5 then shifts the focus to Structured Methods, detailing algorithms that excise complete architectural units to achieve hardware-agnostic, dense-shape acceleration. Following this review, Chapter 6 synthesizes the literature through a cross-regime empirical comparison. It highlights the critical limitations in current evaluation protocols and formally defines the global allocation problem that local scoring methods fail to solve. Finally, Chapter 7 concludes the report, summarizing the core findings and establishing the clear motivation for treating future model compression not as a local ranking task, but as a global configuration search.
