\section{Conclusion}
\vspace{-0.05in}
In this paper, we propose Vector Post-Training Quantization (VPTQ), a novel approach to achieving extremely low-bit quantization of LLMs by Vector Quantization. 
Through the application of Second-Order Optimization, we have formulated the LLM Vector Quantization problem and directed the design of our quantization algorithm. By further refining the weights via Channel-Independent Second-Order Optimization, we have enabled a more granular VQ. 

VPTQ also includes a brief and effective codebook initialization algorithm, which is achieved by decomposing the optimization problem. We have extended VPTQ to support residual and outlier quantization, which not only improves model accuracy but also further compresses the model size. 

Our experimental results demonstrate the effectiveness and efficiency of VPTQ. 
The perplexity of quantized model is reduced by $0.01$-$0.34$ on LLaMA-2, $0.38$-$0.68$ on Mistral-7B, $4.41$-$7.34$ on LLaMA-3 over SOTA at 2-bit, with an average accuracy improvement of  $0.79$-$1.5\%$ on LLaMA-2, $1\%$ on Mistral-7B, $11$-$22\%$ on LLaMA-3  on QA tasks. 
Furthermore, we achieved these results only using $10.4$-$18.6\%$ of the execution time of the quantization algorithm, leading to a $1.6$-$1.8\times$ increase in inference throughput compared to SOTA. 
These results underscore the potential of VPTQ as an efficient and powerful solution for the deployment and inference of LLMs, particularly in resource-constrained settings.
