\chapter{Conclusion}
\label{chap:conclusion}

The three chapters in this report build toward a single observation. Chapter~\ref{chap:foundations} established the formal vocabulary: the Transformer, the efficiency metrics that give compression a concrete meaning, and the genetic search framework that motivates the proposed contribution. Chapter~\ref{chap:neural_network_pruning} defined the pruning problem precisely, surveyed the main criteria from magnitude to second-order approximations, and described the Transformer structures that can be removed without breaking the forward pass. Chapters ch~\ref{chap:related_work_pruning_structured} and ch~\ref{chap:related_work_pruning_unstructured} placed the leading methods in context and identified where each one stops short.

That stopping point is the same across nearly all reviewed methods. Local scoring, whether by weight magnitude, activation statistics, gradient, or curvature, assigns importance to each removable unit before the joint effect of all removals is known. This works at moderate compression. At high compression, the allocation of the budget across layers matters more than the precision of any individual score, and local methods have no mechanism for reasoning about the full configuration. EvoPress makes this concrete: at 37.5\% depth sparsity on Llama-2-7B, methods that score blocks locally reach perplexities above 70, while a search over layer combinations stays at 17.98.

This gap between local scoring and global allocation is the problem the proposed framework is designed to address. The literature reviewed here does not resolve it, but it defines it precisely enough that a search-based approach becomes the natural response.
