\refstepcounter{section}\label{sec:appendix_hps_pythia}

\AppendixPageTitle{Pythia 1.4B Search Hyperparameters}{Full configuration for the joint Drop and Quant evolutionary search.}

\objectivebox{This appendix lists the hyperparameters for the two searches used to compress Pythia 1.4B: Drop for layer removal and Quant for bit-width assignment. It also records the CVT basin construction, EXP3-IX settings, migration policy, restart rules, hill-climbing schedule, fitness definition, and descriptor vectors used to partition the search space.}

\begin{multicols}{2}
\raggedcolumns
\small

\subsection{Algorithm Parameters}

\printbox{Search configuration}{The Drop search selects which layers to remove; the Quant search assigns bit-widths to the retained weights. Each search maintains its own CVT-partitioned population and EXP3-IX bandit. $L=36$ total layers; $k$ is the retained-layer count.}

\begin{cleantable}
\begin{tabularx}{\linewidth}{@{}W{0.43\linewidth}MP{0.215\linewidth}MP{0.215\linewidth}@{}}
\toprule
\textbf{Parameter} & \textbf{Drop} & \textbf{Quant} \\
\midrule
CVT basins $K$ & 5 & 3 \\
Population / basin & 10 & 5 \\
Offspring / gen. & 32 & 32 \\
Generations & $\lfloor k(L-k)/1.5\rfloor$ & 60--80 \\
EXP3-IX $\gamma$ & \multicolumn{2}{c}{0.10} \\
Migration $T_{\mathrm{mig}}$ & \multicolumn{2}{c}{3, ring topology} \\
Stagnation threshold & \multicolumn{2}{c}{5 generations} \\
Periodic restart & every 25 gens & -- \\
Restart strategy & aggressive & gentle \\
Hill-climbing & -- & every 5 gens \\
Fitness & \multicolumn{2}{c}{$-\mathrm{KL}(p_{\mathrm{cmp}}\,\|\,p_{\mathrm{full}})$} \\
CVT samples / Lloyd iterations & \multicolumn{2}{c}{300 / 30} \\
Descriptor dimension & \multicolumn{2}{c}{6} \\
\bottomrule
\end{tabularx}
\end{cleantable}

\subsection{Operators}

\sectioncard{Drop operators \normalfont\small(crossover rate $0.40$)}
\begin{cleantable}
\begin{tabularx}{\linewidth}{@{}W{0.38\linewidth}Y@{}}
\toprule
\textbf{Operator} & \textbf{Effect} \\
\midrule
\code{swap\_one} & Swap one retained/dropped layer pair. \\
\code{swap\_two} & Perform two simultaneous pair swaps. \\
\code{swap\_three} & Perform three simultaneous pair swaps. \\
\code{convert\_to\_block} & Merge scattered drops into one contiguous block. \\
\code{split\_block} & Split a dropped block into two fragments. \\
\code{crossover} & Uniform crossover between two drop masks. \\
\bottomrule
\end{tabularx}
\end{cleantable}

\sectioncard{Quant operators}
\begin{cleantable}
\begin{tabularx}{\linewidth}{@{}W{0.38\linewidth}Y@{}}
\toprule
\textbf{Operator} & \textbf{Effect} \\
\midrule
\code{bit\_swap} & Swap bit-widths of two tensors. \\
\code{bit\_walk\_2} & Budget-neutral $+1/-1$ bit shift. \\
\code{bit\_walk\_3} & Three-way budget-neutral shift. \\
\code{bit\_rebalance} & Redistribute bits according to salience. \\
\code{block\_coord} & Re-optimise one layer block. \\
\code{crossover} & Uniform crossover of bit vectors. \\
\code{hill\_climb} & Greedy single-tensor upgrade. \\
\bottomrule
\end{tabularx}
\end{cleantable}

\columnbreak

\subsection{Descriptors}

\printbox{CVT input}{Each candidate is mapped to a six-dimensional descriptor before basin assignment. Basins are fixed centroids computed from 300 feasible samples with 30 Lloyd iterations. Every descriptor is normalised to $[0,1]$.}

\sectioncard{Drop descriptors}
\begin{cleantable}
\begin{tabularx}{\linewidth}{@{}W{0.43\linewidth}Y@{}}
\toprule
\textbf{Descriptor} & \textbf{Meaning} \\
\midrule
\code{drop\_com} & Centre of mass of dropped sub-block positions. \\
\code{block\_drop\_share} & Fraction of drops that remove both attention and MLP in the same layer. \\
\code{attn\_only\_share} & Share of drop patterns that remove attention only. \\
\code{mlp\_only\_share} & Share of drop patterns that remove MLP only. \\
\code{drop\_clustering} & Clustering score from the variation of gaps between dropped positions. \\
\code{total\_drops\_norm} & Normalised drop count relative to the $2L$ available sub-blocks. \\
\bottomrule
\end{tabularx}
\end{cleantable}

\sectioncard{Quant descriptors}
\begin{cleantable}
\begin{tabularx}{\linewidth}{@{}W{0.43\linewidth}Y@{}}
\toprule
\textbf{Descriptor} & \textbf{Meaning} \\
\midrule
\code{early\_qtr\_share} & Share of total bit budget spent in the first quarter of blocks. \\
\code{late\_qtr\_share} & Share of total bit budget spent in the last quarter of blocks. \\
\code{qkv\_share} & Share of total bit budget spent on QKV input projections. \\
\code{down\_proj\_share} & Down-projection share within the MLP up/down bit budget. \\
\code{long\_hbit\_run} & Longest contiguous run of above-average-bit blocks, divided by the block count. \\
\code{numel\_wtd\_avg} & Parameter-weighted mean bit-width, normalised by the maximum level. \\
\bottomrule
\end{tabularx}
\end{cleantable}

\subsection{Fitness and Restart}

\begin{cleantable}
\begin{tabularx}{\linewidth}{@{}W{0.37\linewidth}Y@{}}
\toprule
\textbf{Item} & \textbf{Value} \\
\midrule
Fitness metric & $-\mathrm{KL}(p_{\mathrm{cmp}}\,\|\,p_{\mathrm{full}})$ \\
Calibration set & Wikitext-2, 128 sequences \\
Shared fitness & Penalty applied to duplicate signatures \\
Drop restart & Reseed from hall of fame plus random candidates \\
Quant restart & Perturb the top-$n$ incumbents \\
\bottomrule
\end{tabularx}
\end{cleantable}

\end{multicols}
\clearpage
