\chapter{Related Work}
\label{chap:related_work_pruning}

\section{Introduction}

Chapter~\ref{chap:neural_network_pruning} established the formal pruning problem, the main sparsity
regimes, the dominant saliency criteria, and the Transformer structures that
can be removed in practice. Against this background, the pruning literature
may be read as a sequence of increasingly structured responses to the same
core problem: how to remove computation while preserving the behavior of the
reference model under explicit deployment constraints.

The literature is not homogeneous: some methods target weight-level sparsity,
others remove executable structures such as heads, FFN channels, or layers,
and others treat pruning as a constrained configuration-search problem under
hardware or latency objectives. The discussion is therefore organized by
pruning mechanism and interpreted with attention to the evaluation regime used
by each paper, since compression level alone is insufficient to compare methods
that optimize different units, use different recovery pipelines, or report
different forms of efficiency.

\input{sections/compression/pruning_methods_pruning_only}
\begingroup
\def\RenderPruningReviewDiscussion{}
\input{sections/compression/pruning_discussion}
\endgroup

\section{Conclusion}

The methods reviewed in this chapter differ in what they score, what they remove, and how much recovery they require, but they share a common limitation at high compression: each method decides locally, before the joint effect of all removals is known. Magnitude and activation-aware methods score individual weights. Gradient and second-order methods improve that score but still apply it one unit at a time. Even dependency-aware structured methods such as LLM-Pruner rank groups independently across layers. The empirical results make the consequence visible: at aggressive depth or sparsity budgets, local ranking degrades rapidly while search over complete configurations does not.
