
\subsection{Hardware-Aware Efficiency Metrics}

Compression is evaluated through several distinct efficiency metrics. Parameter count, memory footprint, arithmetic cost, latency, and throughput capture different aspects of model execution. Hardware-aware evaluation therefore requires separate treatment of storage cost, arithmetic cost, and runtime behavior~\cite{Sze2017EfficientDNN}.

\subsubsection{Parameter Count and Model Size}

Let $\params = \{W^{(l)}, b^{(l)}\}_{l=1}^{L}$ denote the parameters of a network. The total parameter count is definded by Eq.\eqref{eq:parameter_count}:
\begin{equation}
	\label{eq:parameter_count}
	P = \sum_{l=1}^{L} \left( |W^{(l)}| + |b^{(l)}| \right).
\end{equation}
If the parameters are stored with precision $b_w$ bits. Eq.\eqref{eq:model_size} gives the model size in bytes:
\begin{equation}
	\label{eq:model_size}
	M_{\text{model}} = \frac{P\, b_w}{8} \quad \text{bytes}.
\end{equation}
This follows from counting the stored elements and multiplying by the number of bytes used by each element~\cite{PyTorchNumel,PyTorchElementSize}.
This quantity determines whether a model fits within device memory and how much parameter traffic must be sustained during inference. Quantization reduces $b_w$, pruning reduces the number of active parameters, and low-rank factorization reduces the parameterization itself.

For sparse models, storage depends on both the remaining weights and the associated metadata. If $P_{\text{nz}}$ denotes the number of non-zero weights and $M_{\text{meta}}$ the storage needed for indices or masks, the model size is given by \eqref{eq:sparse_model_size}:
\begin{equation}
	\label{eq:sparse_model_size}
	M_{\text{sparse}} = \frac{P_{\text{nz}}\, b_w}{8} + M_{\text{meta}}.
\end{equation}

\subsubsection{MACs, FLOPs, and Arithmetic Intensity}

For a dense matrix multiplication $Y = AB$ with $A \in \reals^{m \times k}$ and $B \in \reals^{k \times n}$, the arithmetic cost is defined by Eq.~\eqref{eq:arithmetic_cost}:
\begin{equation}
	\label{eq:arithmetic_cost}
	\mathrm{MACs}(A,B) = mkn,
	\qquad
	\mathrm{FLOPs}(A,B) \approx 2mkn.
\end{equation}
The distinction is conventional: one multiply-accumulate corresponds to one MAC, while FLOP counting often treats the multiply and the addition as two floating-point operations~\cite{NVIDIAMatrixMultiplication}. In Transformer layers, the feed-forward subnetwork contributes large dense matrix products, while self-attention adds the sequence-dependent $O(n^2 d_{\text{model}})$ cost of score computation and value aggregation.

Arithmetic cost alone is insufficient for hardware analysis. A useful complementary quantity is the arithmetic intensity is defined in Eq.\eqref{eq:arithmetic_intensity}:
\begin{equation}
	\label{eq:arithmetic_intensity}
	I = \frac{\mathrm{FLOPs}}{\mathrm{Bytes\ Moved}},
\end{equation}
which measures how much computation is performed per byte transferred~\cite{Williams2009Roofline,NVIDIAMatrixMultiplication}. Low arithmetic intensity indicates a memory-bound workload, whereas high arithmetic intensity indicates a compute-bound workload.

\subsubsection{Bandwidth, Latency, and Throughput}

Memory bandwidth measures the rate at which data can be transferred between memory and compute units. If $B_{\text{move}}$ bytes are transferred during an operation that takes time $T$, the effective bandwidth demand is calculated by Eq.\eqref{eq:effective_bandwidth}:
\begin{equation}
	\label{eq:effective_bandwidth}
	\mathrm{BW}_{\text{eff}} = \frac{B_{\text{move}}}{T}.
\end{equation}
This follows the standard effective-bandwidth definition used for GPU kernels~\cite{CUDABestPractices}.
When parameter streaming and activation movement dominate execution, bandwidth becomes the primary runtime constraint.

Latency and throughput characterize runtime from complementary viewpoints. Latency is the time required to complete one inference request or one decoding step~\cite{Sze2017EfficientDNN} as shown in Eq.\eqref{eq:latency}:
\begin{equation}
	\label{eq:latency}
	T_{\text{latency}} = T_{\text{compute}} + T_{\text{memory}} + T_{\text{overhead}}.
\end{equation}
Throughput measures completed work per unit time, for example Eq.\eqref{eq:throughput} gives
requests per second or generated tokens per second~\cite{Sze2017EfficientDNN}:
\begin{equation}
	\label{eq:throughput}
	\mathrm{Throughput} = \frac{N_{\text{outputs}}}{T}.
\end{equation}
In large autoregressive models, latency and throughput often follow different scaling trends. Batching can increase throughput while single-request latency remains high. Parameter compression can reduce memory transfer while leaving runtime gains limited when execution is poorly aligned with the hardware schedule.

